---
title: "BST 260 Final Project - COVID-19 Data"
author: "Weixuan Dong, Danny Lee, Bruce Zhou"
format:
  html:
    toc: TRUE
    toc_float: TRUE
    smooth-scroll: TRUE
    math: TRUE
execute:
  echo: false
  warning: false
theme: journal
bibliography: references.bib
---

```{r echo=FALSE}
require(tidyverse)
require(dplyr)
require(lubridate)
require(knitr)
require(ggsignif)
```

```{r echo=FALSE}
rm(list = ls())
load('../data/dat.rda')
load('../data/deaths_all.rda')
```

# Abstract

The COVID-19 pandemic has greatly impacted global health. Especially within the United States, different states experienced significant variations in outcomes over time due to regional differences. This study examines the evolution of COVID-19 virulence and its effects from January 2020 to November 2024, leveraging data from the US Census Bureau and the Centers for Disease Control and Prevention on population estimates, cases, deaths, and hospitalizations. We identified three pandemic waves and assessed state-level performance using metrics such as Case Fatality Rate (CFR), Hospitalization Rate (HR), and Death Proportion (DP).

Our analysis results revealed a pronounced decline in virulence of the COVID-19 virus across three waves, as evidenced by significant reductions in CFR and DP across waves, despite slightly increased transmissibility in later variants. States with stronger healthcare infrastructure and proactive public health policies, such as Hawaii and Massachusetts, generally exhibited lower mortality rates on average. On the contrary, regions with limited resources, like South Dakota and West Virginia, faced higher excess mortality.

Statistical modeling and excess mortality analysis confirm that COVID-19 was the primary driver of increased deaths, and regional disparities were evident. These results underscore the importance of public health systems and targeted interventions in mitigating the impact of pandemics.

# Introduction

The COVID-19 pandemic, first reported in Wuhan in late 2019, is a respiratory disease caused by the severe acute respiratory syndrome coronavirus 2 (SARS-Cov-2) [@shi_2020_an; @covid2021modeling]. The SAR-Cov-2 virus rapidly spread from China to the rest of the world in early 2020, and has arguably constituted the most severe global health crisis the 21st century has seen so far. Beyond its direct impact on human health, the outbreak of COVID-19 has exerted an indelible influence on societies, economies, and the healthcare systems. In the early phase of the COVID-19 pandemic, it was known to cause irreversible pneumonia, high fever, and dry cough [@madabhavi2020covid]. Silvestri et al. (2020) identified four stages of the early variants of the SARS-Cov-2 virus: upper respiratory infection, pneumonia, cytokine storm, and in the worse scenario, death [@stasi2020treatment].

The rapid transmission of the COVID-19 virus was possibly due to its unique characteristic: asymptomatic transmission. Compared with coronaviruses identified previously such as the MERS-Cov, many of the COVID-19 infected patients showed no symptoms during the initial stages of the infection. Despite being asymptomatic, these individuals could unknowingly act as carriers, spreading the virus through aerosols, bodily fluids, or airborne particles [@gao2021systematic]. The feature accelerated the global spread of the COVID-19 virus, and simultaneously hindered the containment efforts led by the healthcare agencies. What's more, as the pandemic progressed and vaccines were developed, the SARS-Cov-2 virus exhibited a high rate of mutation, leading to variants including Alpha, Beta, Delta, and Omicron.  These variants showed differences in their transmissibility, virulence, and the ability to evade the immune system. Because of this, even vaccines, which used to be taken as the ultimate measure to end any pandemic, proved to fall short of people's expectation in controlling this worldwide crisis [@vasireddy2021review]. As a result, the majority of the governments announced emergency measures such as mask mandates, social distancing, regular sterilization, and even full lock-down of cities [@koh2020covid].

Especially in the United States, the COVID-19 pandemic has been complicated and exacerbated by various factors [@omer2020covid; @covid2021modeling]. Given the high degree of autonomy of each state in the US, significant variations could be seen across states due to discrepancies in regional public health policies, demographics, and economic development. Previous studies have suggested that the ethnic minorities, middle- and low-income populations, and rural communities were particularly more vulnerable to this emergency. These populations often suffer from pre-existing health inequities and consequently disproportional rates of mortality were reported [@alcendor2020racial; @wilder2021disproportionate].

The US Centers for Disease Control and Prevention (CDC) provides an excellent platform for collecting and sharing COVID-19 related data. These data enable us to study the progression and impact of COVID-19 throughout the pandemic period. We can expect that different states might have responded differently: adopting distinct strategies for containment and mitigation. Additionally, as the COVID-19 virus mutated, multiple waves could be identified, which were driven by seasonal changes, vaccine effectiveness, and the fatality rate. Understanding these patterns is vital to assessing the evolution of the SARS-Cov-2 virus, regional differences, and the virulence over time.

In this project, we examined the COVID-19 data provided by the CDC COVID-19 Response, including COVID-19 cases, COVID-related deaths, and hospitalization from Jan 25th, 2020 to Nov 30th, 2024. Using data analysis and visualization, this study aims to identify major waves of the COVID-19 pandemic in the US, and answer the following questions 1) which states generally responded to the pandemic better, 2) how the virulence of the SARS-Cov-2 virus evolved and examine regional differences and seasonal changes in the transmissibility of the virus. In addition, we also seek to estimate weekly excess mortality for each state during the pandemic period and investigate whether regional differences could be observed and COVID-19 was the driving factor. We hypothesize that states along the West and East Coasts would generally perform better during the pandemic due to stronger healthcare infrastructure. We also hypothesize that the COVID-19 virus became less virulent over time as viruses typically do not exhibit both high virulence and high transmissibility.

# Methods

First, we constructed a population matrix using two datasets from the United States Census Bureau (USCB). For the years 2020 and 2021, we used observed annual population data for all states, which is available here <https://www.census.gov/data/developers/data-sets/popest-popproj/popest.html>. For the years 2022 and 2023, we used the Annual National Population Estimates data, which includes both national and state-level population estimates. It can be assessed through: <https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html>. These were combined to form a comprehensive population matrix. Finally, we applied a simple linear regression model to predict state-level population figures for 2024 based on trends observed in the preceding years, completing the dataset for 2020-2024. Our extrapolation stands under the assumption that the trends observed in preceding years would similarly be observed in 2024. This potentially poses some limitations of using linear regression model as there could be possible non-linear dynamics for year 2024 such as natural disasters in some states.

Next, we used the dataset from the Centers for Disease Control and Prevention (CDC), containing COVID-19 cases, COVID-related deaths, hospitalization, and vaccination data from January 25th, 2020 to November 30th, 2024. Data wrangling was conducted to store them as either weekly or daily data, enabling more interpretable analysis going forward. The corresponding wrangling scripts could be found in `code/wrangle.R`.

Finally, we constructed a deaths data for all causes for year 2020-2024 as this was salient in estimating the excess mortalities of the COVID-19 in the period of our interest. We first used the Weekly counts of deaths by Jurisdiction and Age data provided by National Center for Health Statistics (NCHS) to extract the historical data of “all-cause deaths” from years 2015-2019. Then, we fit a simple linear regression model per state predicting all-cause deaths by utilizing yearly trends and weekly trends for extrapolating 2020-2023 mortality by all causes (not limited to deaths caused by COVID-19). During the process of extrapolation, we fit other linear models by adding more complex seasonality terms such as sine / cosine of week or factorized weeks, but just simply fitting the simple linear model provided the most parsimonious and interpretable model. Similar to the estimation of the total population in 2024, this estimation is also based on the assumption that the all-cause mortality is linear on yearly trends and weekly trends. 

Our analysis of COVID-19 virulence uses several analytic metrics, primarily **Case Fatality Rate (CFR)**, **Hospitalization Rate (HR)**, and **Death Proportion (DP)**. These metrics are defined as follows:

$$
\text{CFR} = \frac{\text{Total Deaths}}{\text{Total Cases}} \times 100,000
$$

$$
\text{HR} = \frac{\text{Total Hospitalizations}}{\text{Total Cases}} \times 100,000
$$

$$
\text{DP} = \frac{\text{Total Deaths}}{\text{Total Population}} \times 100,000
$$

Also, to compare the three metrics stated above across waves, we used Wilcoxon rank-sum test for non-parametric comparisons. It assesses whether the distribution of a particular variable is significantly different between the two groups when the data does not meet the assumptions of normality required for a t-test. We set the null hypothesis to “the two groups come from populations with the same distribution”, and the alternative as “the two groups come from populations with different distributions”. In our research project, this would be interpreted as testing whether COVID-19 metrics (e.g. CFR, HR, DP) during Wave 1 vs. Wave 2 show significant difference between the two waves.

Our excess mortality analysis uses the definition Excess Mortality = Observed Deaths – Estimated Deaths to calculate excess mortality. Our methodology is that after fitting the linear model with outcome variable being excess deaths and predictor being covid deaths, if the summary statistics show high R-squared or significant p-value, we conclude that covid deaths is a strong predictor, advocating the COVID-19 was indeed the culprit of the excess deaths in US states. We also visually check whether the COVID-19 is the cause of excess mortalities by drawing a weekly excess mortality plot with COVID-19 caused deaths plotted on top of it. 

Our methods is robust since it divides the pandemic into specific waves, thus acknowledges the temporal and dynamic nature of the pandemic, allowing for contextual insights. 

# Analyses

## Q1: Pandemic period from Jan 2020 to Nov 2024

The figure below displays the COVID-19 cases, deaths, and hospitalization rate per 100,000 people in each region of the US from 2020 to 2024.

```{r trend plots, echo=FALSE, warning=FALSE}
#| fig-width: 12
#| fig-height: 8

# Define custom labels for the facets
dat.labs <- c("cases" = "COVID-19 Cases", "deaths" = "Deaths", "hosp" = "Hospitalizations")

# Apply to the ggplot code
dat %>%
  pivot_longer(cols = c("cases", "deaths", "hosp"), names_to = "outcome", values_to = "count") %>%
  mutate(rate = count * 100000 / population) %>%
  ggplot(aes(x = date, y = rate, group = state, color = region_name)) +
  geom_line(linewidth = 1, alpha = 0.7) + # Thicker lines for better visibility
  facet_wrap(~outcome, ncol = 1, scales = "free", labeller = labeller(outcome = as_labeller(dat.labs))) +
  scale_x_date(
    breaks = "6 months", # Show a date label every 6 months
    date_labels = "%b %Y" # Format the labels as "Month Year" (e.g., Jan 2021)
  ) +
  theme_minimal(base_size = 14) + # Use a cleaner theme with larger base font
  theme(
    strip.text = element_text(face = "bold", size = 14), # Bold facet labels
    legend.position = "bottom", # Move legend to bottom for better space usage
    legend.title = element_text(face = "bold"), # Bold legend title
    title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"), # Bold axis titles
    axis.text.y = element_text(size = 10), # Slightly larger axis text for readability
    axis.text.x = element_text(size = 8, angle = 10),
    panel.grid.minor = element_blank() # Remove minor grid lines for a cleaner look
  ) +
  labs(
    x = "Date",
    y = "Rate (per 100,000 people)",
    color = "Region",
    title = "Rates of Cases, Deaths, and Hospitalizations Over Time (State Level)",
    subtitle = "Data displayed per 100,000 people by region"
  )

# ggsave(filename = "../output/covidwaves.pdf",plot = p1, dpi = 300, width = 14, height = 8, units = "in")
```

As can be seen from the trend plot which depicts the rates of COVID-19 cases, deaths, and hospitalizations from 01/25/2021 to 11/30/2024, there seems to be 3 waves of COVID-19 spread. The first one occurs from Oct 2020 to Apr 2021, where a peak in cases, deaths, and hospitalizations can be observed. Likewise, it appears that there was another wave between Jul 2021 to early Dec 2021. Lastly, there was a wave from early Dec 2021 to Apr 2022.

**First Wave**

This period marked the first significant rise in COVID-19 rates, with sharp increases across all metrics: cases, deaths, and hospitalizations. The wave coincides with the colder months when indoor gatherings increased and may also reflect the impact of delayed public health measures or variants circulating during this time.

**Second Wave**

The summer and early winter of 2021 saw a resurgence in COVID-19 activity, driven largely by the spread of the Delta variant. This wave demonstrates a rapid increase in hospitalizations and deaths following the initial rise in cases. Despite vaccination efforts, breakthrough infections and the highly contagious nature of Delta played significant roles in this period.

**Third Wave**

The final and most intense wave depicted occurred during the winter of 2021-2022, fueled by the Omicron variant. While cases skyrocketed to unprecedented levels due to Omicron's transmissibility, deaths and hospitalizations rose less dramatically than in previous waves, likely reflecting the variant's comparatively milder effects and increased vaccination coverage. However, the healthcare system faced immense strain due to the sheer number of cases.

## Q2: Deaths rate by state during each time period

```{r separately save each time period}
wave1dat <- dat %>%
  filter(date >= make_date(2020, 10, 1), date <= make_date(2021, 4, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  ) %>%
  ungroup()

wave2dat <- dat %>%
  filter(date >= make_date(2021, 7, 1), date <= make_date(2021, 12, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  ) %>%
  ungroup()

wave3dat <- dat %>%
  filter(date >= make_date(2021, 12, 2), date <= make_date(2022, 4, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  ) %>%
  ungroup()

# write_csv(wave1dat,file = "../output/wave1dat.csv")
# write_csv(wave2dat,file = "../output/wave2dat.csv")
# write_csv(wave3dat,file = "../output/wave3dat.csv")
```

From Oct 1st 2020 to Apr 1st 2021, overall this wave was arguably the most severe wave among all 3. The highest death rate per 100,000 people was reported in South Dakota (SD), with an average of 188.91 deaths among 100,000 people. Following South Dakota were North Dakota (177.97), Oklahoma (175.32), and Arkansas (150.17). There were also states which managed to contain the death rate below 50 per 100,000. For example, the lowest death rate was reported in Hawaii, which was only 9.12 deaths on average per 100,000 people, followed by Vermont (11.18), Alaska (26.48), Washington (39.58), Oregon (42.04), Puerto Rico (43.66), and Maine (47.19).

From July 1st 2021 to Dec 1st 2021, the COVID-19 variant seemed to be less virulent. The highest death rate per 100,000 people was reported in West Virginia (WV), with an average of 113.13 deaths among 100,000 people. Following West Virginia were Florida  (111.62), Montana (111.30), and Wyoming (109.88). There were also states which managed to contain the death rate below 40 per 100,000. For example, the lowest death rate was reported in Washington D.C., which was only 12.09 deaths on average per 100,000 people, followed by Vermont (12.39), Connecticut (12.39), New York (17.04), and Hawaii (36.83).

For the third wave, the COVID-19 variant during the period seemed to grow more infectious and deadly. The highest death rate per 100,000 people was reported in West Virginia (WV) again, with an average of 103.73 deaths among 100,000 people. Following West Virginia were Ohio (102.05), New Mexico (96.78), and Kentucky (93.82). There were also states which managed to contain the death rate below 40 per 100,000. For example, the lowest death rate was reported in Hawaii, which was only 20.76 deaths on average per 100,000 people, followed by Alaska (27.69), Puerto Rico (28.21), Vermont (29.85), and Utah (33.22). Figures can be found in **Supplementary Methods: Death rate of each state (3 waves)**.

We also compared the performance of each state across all 3 waves by taking the average of their death rate per 100,000 people. Results show that Oklahoma had the highest average death rate per 100,000, amounting to 120.00 deaths per 100,000 people, followed by West Virginia (115.33), Mississippi (109), Kentucky (107.42), Tennessee (106.92). There were also states which kept a relatively low death rate across waves. Namely, Vermont had an average death rate of only 17.81 per 100,000. Similarly, Hawaii had an average of 22.24 per 100,000, followed by Puerto Rico (30.94) New York (37.10), and Washington (40.07).

Generally, it can be seen that states in the northeastern part or the west coast of the US performed better than those in the middle or west US. It can also be observed that less populated states such as Vermont and Washington had lower death rates. But exceptions are also evident, for New York is rather populated but it was among the states with the lowest average death rates.

```{r avg death rate across 3 waves}
#| fig-width: 12
#| fig-height: 8

wave1dat$wave <- "Wave 1"
wave2dat$wave <- "Wave 2"
wave3dat$wave <- "Wave 3"
allWaves <- rbind(wave1dat, wave2dat, wave3dat)

allWaves %>%
  group_by(state) %>%
  summarise(avg_death_rate = mean(death_rate, na.rm = T)) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(state, avg_death_rate), y = avg_death_rate, fill = avg_death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient2(low = "#FFE5B4", mid = "#FFA500", high = "#800000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    title = element_text(face = "bold"),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Average Death Rate (per 100,000)",
    fill = "Average Death Rate",
    title = "Average COVID-19 Death Rates by State across 3 Waves"
  )
```

## Q3: COVID-19 became less or more virulent across different periods

To evaluate if COVID-19 became less or more virulent across different time periods, we will employ three key metrics: **cases fatality rate (CFR)**, **hospitalization rate (HR)**, and **death per population (DP)**. The CFR is defined as deaths per case, HR is defined as hospitalizations per case, and DP is defined as deaths per population. In the following analyses, we will analyze these key metrics with 
tables and visualizations.

```{r create dataset with wave indicator}
## Create dataset with wave indicator ##
pooled.dat <- dat %>%
  mutate(wave = case_when(
    date >= make_date(2020, 10, 1) & date <= make_date(2021, 4, 1) ~ "Wave 1",
    date >= make_date(2021, 7, 1) & date <= make_date(2021, 12, 1) ~ "Wave 2",
    date >= make_date(2021, 12, 2) & date <= make_date(2022, 4, 1) ~ "Wave 3",
    .default = NA
  )) %>%
  filter(!is.na(wave)) %>%
  group_by(wave, state) %>%
  summarise("Total Cases" = sum(cases, na.rm = T),
            "Total Deaths" = sum(deaths, na.rm = T),
            "Total Hospitalization" = sum(hosp, na.rm = T),
            "Total Population" = mean(population, na.rm = T), .groups = "drop")

## Calculate the key metrics ##
metric.dat <- pooled.dat %>%
  mutate(CFR = `Total Deaths` * 100000 / `Total Cases`,
         HR = `Total Hospitalization` * 100000 / `Total Cases`,
         DP = `Total Deaths` * 100000 / `Total Population`)
```

### Cases fatality rate, hospitalization rate, and death per population

```{r}
# Define custom labels for the facets
dat.labs <- c("CFR" = "Cases Fatality Rate", "HR" = "Hospitalization Rate", "DP" = "Death per Population")

library(scales)

pooled.dat %>%
  group_by(wave) %>%
  summarise("Total Cases" = sum(`Total Cases`, na.rm = TRUE),
            "Total Deaths" = sum(`Total Deaths`, na.rm = TRUE),
            "Total Hospitalization" = sum(`Total Hospitalization`, na.rm = TRUE),
            "Total Population" = sum(`Total Population`, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(CFR = `Total Deaths` * 100000 / `Total Cases`,
         HR = `Total Hospitalization` * 100000 / `Total Cases`,
         DP = `Total Deaths` * 100000 / `Total Population`) %>%
  mutate(
    `Total Cases` = comma(`Total Cases`),
    `Total Deaths` = comma(`Total Deaths`),
    `Total Hospitalization` = comma(`Total Hospitalization`),
    `Total Population` = comma(`Total Population`),
    CFR = comma(CFR, accuracy = 0.01),
    HR = comma(HR, accuracy = 0.01),
    DP = comma(DP, accuracy = 0.01)
  ) %>%
  rename(Wave = wave) %>%
  kable("simple")

# pooled.dat %>%
#   group_by(wave) %>%
#   summarise("Total Cases" = sum(`Total Cases`, na.rm = T),
#             "Total Deaths" = sum(`Total Deaths`, na.rm = T),
#             "Total Hospitalization" = sum(`Total Hospitalization`, na.rm = T),
#             "Total Population" = sum(`Total Population`, na.rm = T)) %>%
#   ungroup() %>%
#   mutate(CFR = `Total Deaths` * 100000 / `Total Cases`,
#          HR = `Total Hospitalization` * 100000 / `Total Cases`,
#          DP = `Total Deaths` * 100000 / `Total Population`) %>%
#   pivot_longer(
#     cols = c(CFR, HR, DP),
#     names_to = "metric",
#     values_to = "rate"
#   ) %>%
#   ggplot(aes(x = wave, y = rate, fill = metric)) +
#   geom_col(show.legend = FALSE) +
#   facet_wrap(
#     ~ metric,
#     scales = "free",
#     labeller = labeller(metric = dat.labs)
#   ) +
#   scale_fill_brewer(palette = "Set2") +
#   theme_minimal() +
#   theme(
#     strip.text = element_text(size = 12, face = "bold"),
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
#     axis.title = element_text(size = 12, face = "bold")
#   ) +
#   labs(
#     title = "COVID-19 Aggregated Virulence Metrics by Wave",
#     x = "Wave",
#     y = "Rate (per 100,000)"
#   )
```

**Wave 1**

During Wave 1, the United States experienced a total of 22,403,089 cases, with 348,013 deaths and 1,714,882 hospitalizations. The total population during this period was 334,955,562. Key metrics calculated include:

* Case Fatality Rate (CFR): 1,553.42 per 100,000 cases.
* Hospitalization Rate (HR): 7,654.67 per 100,000 cases.
* Death Proportion (DP): 103.90 deaths per 100,000 population.

**Wave 2**

Wave 2 saw a decrease in total cases and deaths compared to Wave 1, with 14,304,772 cases, 194,568 deaths, and 1,122,040 hospitalizations. The total population for this wave was 335,157,329. Metrics for this wave include:

* Case Fatality Rate (CFR): 1,360.16 per 100,000 cases.
* Hospitalization Rate (HR): 7,843.82 per 100,000 cases.
* Death Proportion (DP): 58.05 deaths per 100,000 population.

**Wave 3**

Wave 3 reported the highest total cases at 30,709,203, but a similar total death count compared to Wave 2, with 194,326 deaths and 1,183,261 hospitalizations. The total population during this wave was 336,099,114. Metrics calculated for Wave 3 include:

* Case Fatality Rate (CFR): 632.79 per 100,000 cases.
* Hospitalization Rate (HR): 3,853.12 per 100,000 cases.
* Death Proportion (DP): 57.82 deaths per 100,000 population.

For each wave, we computed the CFR, HR, and the DP. It can be observed that total cases peaked in Wave 3, while total deaths and hospitalizations decreased relative to Wave 1. The CFR showed a pronounced decline across waves, decreasing from 1,553.42 in Wave 1 to 632.79 in Wave 3 (59.26%). The HR also declined significantly by Wave 3, from 7,654.67 in Wave 1 to 3,853.12 in Wave 3. Similarly, the DP decreased steadily across the waves, dropping from 103.90 in Wave 1 to 57.82 in Wave 3.

We conducted pairwise statistical tests on the CFR, HR, and DP metrics to evaluate differences across the pandemic waves. Since these metrics did not follow a normal distribution, we utilized the Wilcoxon rank-sum test, a non-parametric alternative to the independent t-test. The analysis revealed a statistically significant decline in the CFR from Wave 1 to Wave 2, with a p-value of less than 0.001. Similarly, the CFR dropped significantly from Wave 2 to Wave 3. When comparing Wave 1 to Wave 3, the difference was also statistically significant at the 0.05 significance level.

For the HR, there was no statistically significant difference between Wave 1 and Wave 2, as indicated by a p-value of 0.757. However, a notable decline in HR was observed in Wave 3 compared to both Wave 1 and Wave 2, with p-values of less than 0.001. These findings demonstrate a clear downward trend in the CFR across all three waves, while the HR remained relatively stable between the first two waves before showing a significant reduction in the third wave.

Finally for the DP, there was a statistically significant difference between Wave 1 and Wave 2, with a p-value of less than 0.001. A highly significant difference was also observed between Wave 1 and Wave 3, indicating a sharp decline in DP over time. However, the comparison between Wave 2 and Wave 3 revealed no statistically significant difference, as evidenced by a p-value of 0.647. Figures can be found in **Supplementary Methods: Pair-wise comparison of the CFR, HR, and DP using statistical tests**.

## Q4: Estimate excess mortality for each week for each state. Check if excess mortality can be explained by COVID-19

```{r}
#| fig-width: 12
#| fig-height: 8

library(tidyr)
library(purrr)
library(broom)
library(ggplot2)

dat <- dat %>% 
  rename(covid_deaths = deaths) # renaming for clarity

##############################
# 1. Prepare historical data (2015-2019)
##############################

# Filter historical period
historical_data <- deaths %>%
  filter(mmwr_year >= 2015 & mmwr_year <= 2019) 

##############################
# 2. Fit Linear Models to Estimate Baseline
##############################

# We'll fit a simple linear model per state predicting all_cause_deaths
# from mmwr_year and mmwr_week. You could add more complex seasonality terms,
# such as using sin/cos of week or factorized weeks, but here's a simple approach.

models <- historical_data %>%
  group_by(state_name) %>%
  do(
    model = lm(all_cause_deaths ~ mmwr_year + mmwr_week, data = .)
  )

##############################
# 3. Predicting Expected Deaths for 2020–2024
##############################

# Filter data for prediction period (2020+)
future_data <- deaths %>%
  filter(mmwr_year >= 2020)

# Generate predictions by applying each state's model to its future_data
future_predictions <- future_data %>%
  group_by(state_name) %>%
  nest() %>%
  left_join(models, by = "state_name") %>%
  mutate(predictions = map2(data, model, ~augment(.y, newdata = .x))) %>%
  unnest(predictions) %>%
  ungroup()

# Now `future_predictions` contains:
# - all original columns from `future_data`
# - .fitted: predicted (expected) deaths
# - .se.fit, .resid, etc., if needed

##############################
# 4. Calculate Excess Mortality
##############################

future_predictions <- future_predictions %>%
  mutate(excess_deaths = all_cause_deaths - .fitted)

##############################
# 5. Merge With COVID-19 Deaths
##############################

# Ensure `dat` also has mmwr_year, mmwr_week, and state_name for the same weeks
analysis_data <- future_predictions %>%
  left_join(dat %>% select(state_name, mmwr_year, mmwr_week, covid_deaths),
            by = c("state_name", "mmwr_year", "mmwr_week"))

# Now `analysis_data` has: all_cause_deaths, expected deaths (.fitted), excess_deaths, and covid_deaths

##############################
# 6. Visualizations
##############################

analysis_data <- analysis_data %>%
  filter(!state_name %in% c("New York City", "United States")) %>%
  # Approximate a date from mmwr_year and mmwr_week:
  mutate(approx_date = as.Date(paste(mmwr_year, mmwr_week, 1, sep="-"), "%Y-%U-%u"))

# Facet by state to get a quick overview:
p_all_states <- analysis_data %>%
  ggplot(aes(x = approx_date)) +
  geom_line(aes(y = excess_deaths, color = "Excess Deaths")) +
  geom_line(aes(y = covid_deaths, color = "COVID-19 Deaths")) +
  facet_wrap(~ state_name, scales = "free_y") +
  labs(title = "Excess vs. COVID-19 Deaths by State",
       x = "Date", 
       y = "Number of Deaths",
       color = "") +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  )

print(p_all_states)

# If you have many states and the plot looks crowded, consider plotting one state at a time or saving to a file:
# Example of saving one plot per state:
# unique_states <- unique(analysis_data$state_name)
# for (st in unique_states) {
#   p_state <- analysis_data %>%
#     filter(state_name == st) %>%
#     ggplot(aes(x = approx_date)) +
#     geom_line(aes(y = excess_deaths, color = "Excess Deaths")) +
#     geom_line(aes(y = covid_deaths, color = "COVID-19 Deaths")) +
#     labs(title = paste("Excess vs. COVID-19 Deaths:", st),
#          x = "Date",
#          y = "Number of Deaths",
#          color = "") +
#     theme_minimal()
#   
#   ggsave(filename = paste0("excess_vs_covid_", st, ".png"), plot = p_state, width = 8, height = 5)
# }


##############################
# 8. Interpret Results
##############################

# If the summary of lm_result shows covid_deaths is a strong predictor (high R-squared, significant p-value),
# this indicates COVID-19 deaths explain a large portion of excess mortality.
# If correlation is high and the slope ~ 1, it suggests a strong direct relationship.

##############################
# Additional Considerations:
##############################

### might want to incorporate seasonality more explicitly (e.g., factor(mmwr_week)) in the baseline model.
### Try more advanced models (ARIMA, GAM) for better baseline estimation?
```

We are first using 2015-2019 US mortality data grouped by states to estimate the 2020-2023 mortality by all causes (not limited to deaths caused by Covid-19). Then we calculate excess mortality by subtracting the estimated deaths from observed deaths of 2020-2023. This is depicted as blue line in the plot. Now, we plot the deaths which the culprit is Covid-19 via red line to see whether the excess mortality is well explained by Covid-19. We can see that in most of the states, COVId-19 is indeed the main culprit of excess mortality from 2020-2023. However, some states such as New Hampshire, Oregon, Puerto Rico, and Washington do not have a similar trend of lines between COVID-19 deaths and Excess deaths, meaning COVID-19 does not explain the excess mortality that well.

Second, we conducted additional analysis of fitting the linear model with COVID-19 deaths being the covariate and excess deaths being the outcome variable. The results of the model advocated the findings that we have roughly found from the plot. The R-squared values of each state's model showed that New Hampshire(0.236), Oregon(0.301), Puerto Rico(0.194), and Washington(0.257)'s respective R-squared values were all low compared to other states. Some states such as New Jersey, North Dakota, and South Dakota showed very high R-squared values of 0.851, 0.849, and 0.857 respectively which corresponded to the alignment of Covid caused deaths and excess deaths in the plot. Detailed regression ouputs can be found in **Supplementary Methods: Regression (fitted) table of all states in the U.S.**.

Overall, we can conclude that COVID-19 fairly explains the excess mortality in the United States during 2020-2023 well!

## Q5. For each period compute the excess mortality rates by state. Describe which states did better or worse during the different periods.

```{r}
#| fig-width: 12
#| fig-height: 8
analysis_data <- analysis_data |>
  left_join(dat |> distinct(state_name, population), by = "state_name", relationship = "many-to-many")

# Define wave periods
wave_periods <- list(
  Wave1 = c("2020-10-01", "2021-04-01"),
  Wave2 = c("2021-07-01", "2021-12-01"),
  Wave3 = c("2021-12-02", "2022-04-01")
)

# Function to compute excess mortality rate by wave
compute_excess_rate <- function(data, start_date, end_date) {
  data |>
    filter(approx_date >= as.Date(start_date) & approx_date <= as.Date(end_date)) |>
    group_by(state_name) |>
    summarise(
      total_excess_deaths = sum(excess_deaths, na.rm = TRUE),
      population = mean(population, na.rm = TRUE)
    ) |>
    mutate(excess_death_rate = total_excess_deaths * 100000 / population) |>
    ungroup()
}

# Calculate excess mortality rates for each wave
wave1_excess <- compute_excess_rate(analysis_data, wave_periods$Wave1[1], wave_periods$Wave1[2])
wave2_excess <- compute_excess_rate(analysis_data, wave_periods$Wave2[1], wave_periods$Wave2[2])
wave3_excess <- compute_excess_rate(analysis_data, wave_periods$Wave3[1], wave_periods$Wave3[2])

# Write results to CSV files
#write_csv(wave1_excess, file = "../output/wave1_excess.csv")
#write_csv(wave2_excess, file = "../output/wave2_excess.csv")
#write_csv(wave3_excess, file = "../output/wave3_excess.csv")

# Visualization for excess mortality rates for all states across 3 waves
# Combine all waves into one data frame with a "Wave" column
wave1_excess$Wave <- "Wave 1"
wave2_excess$Wave <- "Wave 2"
wave3_excess$Wave <- "Wave 3"

all_excess <- bind_rows(wave1_excess, wave2_excess, wave3_excess)

# Create a factor for Wave to control the order in the plot if desired
all_excess$Wave <- factor(all_excess$Wave, levels = c("Wave 1", "Wave 2", "Wave 3"))

# Plot: Bar plot of excess death rates for all states, faceted by Wave
ggplot(all_excess, aes(x = reorder(state_name, excess_death_rate), y = excess_death_rate, fill = excess_death_rate)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~Wave, scales = "free") +
  scale_fill_gradient(low = "#FFDDC1", high = "#AE0000") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(face = "bold"),
    strip.text = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 7),
    title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  labs(
    y = "Excess Mortality Rate (per 100,000)",
    fill = "Excess Mortality",
    title = "Excess Mortality Rates by State Across 3 Waves",
    subtitle = "Each facet shows a different wave period"
  )
```

```{r}
# Top and bottom 5 for each wave
top5_wave1 <- wave1_excess %>% arrange(desc(excess_death_rate)) %>% slice_head(n=5) %>% mutate(Rank = row_number())
bottom5_wave1 <- wave1_excess %>% arrange(excess_death_rate) %>% slice_head(n=5) %>% mutate(Rank = 53 - row_number())

top5_wave2 <- wave2_excess %>% arrange(desc(excess_death_rate)) %>% slice_head(n=5) %>% mutate(Rank = row_number())
bottom5_wave2 <- wave2_excess %>% arrange(excess_death_rate) %>% slice_head(n=5) %>% mutate(Rank = 53 - row_number())

top5_wave3 <- wave3_excess %>% arrange(desc(excess_death_rate)) %>% slice_head(n=5) %>% mutate(Rank = row_number())
bottom5_wave3 <- wave3_excess %>% arrange(excess_death_rate) %>% slice_head(n=5) %>% mutate(Rank = 53 - row_number())

# Combine all data
combined_data <- bind_rows(
  top5_wave1,
  bottom5_wave1,
  top5_wave2,
  bottom5_wave2,
  top5_wave3,
  bottom5_wave3
)
combined_data <- combined_data |>
  rename(
    State = state_name,
    "Excess Mortality" = total_excess_deaths,
    " Excess Mortality Rate per 100,000" = excess_death_rate,
    Population = population
) 

# View the combined data
kable(combined_data, digit = 3, align = "c")
```

We calculated each state's excess mortality during three waves by estimating mortality based on the linear model fitted with the 2015-2019 mortality data and subtracting this from observed deaths. 

For the first wave(Oct 1, 2020 – Apr 1, 2021): South Dakota, Mississippi, Oklahoma, Arkansas, and New Mexico had highest excess mortality rates. Hawaii, Washington, New Hampshire, Maine, and Massachusetts had lowest excess mortality rates. South Dakota and Mississippi showed significant excess mortality, because in these areas, they have limited healthcare access. And, Hawaii and Washington showed lower rates which they had early public health interventions and stronger healthcare systems.

For the second wave(Jul 1, 2021 – Dec 1, 2021): Starting from wave 2, the Covid_19 vaccine came out. West Virginia, Alaska, Montana, Wyoming, and Mississippi had the highest rates. New Jersey, Massachusetts, Maryland, Connecticut, and New York had the lowest rates. West Virginia and Mississippi had highest excess mortality rates because they have limited healthcare infrastructure and lower vaccine rate. States like New Jersey and Massachusetts have access to various healthcare resources where higher vaccination rates and effective policy which contributed to their lower rates.

For the third wave(Dec 2, 2021 – Apr 1, 2022): Fueled by the Omicron variant, West Virginia, New Mexico, Oklahoma, Kentucky, and Arizona had the highest excess mortality rates. States with the lowest rates were Hawaii, New York, Utah, Massachusetts, and Washington. West Virginia and Oklahoma continued to have highest excess mortality. On the other hand, Hawaii and Massachusetts consistently performed well due to their effective public health policies.

Overall, the results above shows that COVID-19 explains excess mortality in the United States during 3 waves and also how vaccination and public health systems are involved in the prevention of excess deaths during the pandemic.

# Discussion

In this study, we analyzed COVID-19 data across the United States from January 2020 to November 2024, focusing on evaluating state-level performance during the pandemic. Our approach combined statistical modeling, excess mortality estimation, and comparisons across three defined pandemic waves. By integrating case fatality rates (CFR), hospitalization rates (HR), and death proportions (DP), we separate the entire periods into 3 waves where a peak in cases, deaths, and hospitalizations can be observed. The first one occurs from Oct 2020 to Apr 2021, there are other waves between Jul 2021 to early Dec 2021 and between from early Dec 2021 to Apr 2022. We used pre-pandemic mortality trends (2015–2019) to estimate expected deaths if no Covid-19 happened, which we provided a standard to identify excess mortality caused by Covid-19.

Our analysis showed big differences between states and waves. In Wave 1, excess mortality was highest in states like South Dakota and Mississippi where structural issues like limited healthcare resources made the virus worse. Hawaii and Washington had much lower excess mortality likely due to strong healthcare systems and early interventions. Under Wave 2, mortality metrics improve in many states due to Covid-19 vaccines came out. But challenges remained especially in rural states like West Virginia and Alaska where healthcare access and vaccine distribution were limited. Massachusetts and New Jersey did well due to good public health policies and higher vaccination rates. Wave 3 showed the impact of the super spreader Omicron variant. While cases skyrocketed to record levels, mortality and hospitalization rates didn’t rise as much, due to the protection of vaccines. But West Virginia and New Mexico still had high excess mortality, highlighting the persistent healthcare infrastructure disparities. Hawaii and New York had lower mortality rates, due to sustained public health policies. Our results show that public health policies, vaccination campaigns and healthcare infrastructure matter. States with strong and sustained interventions did well across all waves. States with structural challenges like limited healthcare capacity and delayed public health response had more mortality.

While our analysis provided valuable insights, it also has limitations. First, our excess mortality estimates used linear trends from historical data which may not capture non-linear or seasonal patterns. Second, the indirect effects of the pandemic like delayed medical care or socioeconomic stressors were not modeled but likely contributed to excess mortality in some states. Third, variations in reporting practices across states may have introduced inconsistencies in the data.

When doing similar analysis in the future, it’s necessary to consider the assumptions behind the mortality trend extrapolations especially when there are big disruptions like the pandemic. Sensitivity analysis and model diagnostics are crucial to check the results. Future studies should also look into the long term effects of the pandemic including the indirect effects of delayed care and economic stressors and how healthcare inequities contributed to the observed disparities.

# Supplementary Methods

## Death rate of each state (3 waves)
```{r death rate by state}
#| fig-width: 8
#| fig-height: 6

wave1dat %>%
  ggplot(aes(x = reorder(state, death_rate), y = death_rate, fill = death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient(low = "#FFA5A5", high = "#AA0000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Death Rate (per 100,000)",
    fill = "Death Rate",
    title = "Cumulative COVID-19 Death Rates by State",
    subtitle = "Data from Oct 1st 2020 to Apr 1st 2021"
  )

wave2dat %>%
  ggplot(aes(x = reorder(state, death_rate), y = death_rate, fill = death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient(low = "#FFA5A5", high = "#AA0000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Death Rate (per 100,000)",
    fill = "Death Rate",
    title = "Cumulative COVID-19 Death Rates by State",
    subtitle = "Data from Jul 1st 2021 to Dec 1st 2021"
  )

wave3dat %>%
  ggplot(aes(x = reorder(state, death_rate), y = death_rate, fill = death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient(low = "#FFA5A5", high = "#AA0000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Death Rate (per 100,000)",
    fill = "Death Rate",
    title = "Cumulative COVID-19 Death Rates by State",
    subtitle = "Data from Dec 2nd 2021 to Apr 1st 2022"
  )
```

## Pair-wise comparison of the CFR, HR, and DP using statistical tests

```{r warning=FALSE}
#| fig-width: 8
#| fig-height: 6

## Plot the statistical significance ##
metric.dat %>%
  ggplot(aes(x = wave, y = CFR, fill = wave)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(
      c("Wave 1", "Wave 2"),
      c("Wave 1", "Wave 3"),
      c("Wave 2", "Wave 3")
    ),
    y_position = c(2500, 2700, 3000), # Adjust for clear separation
    test = wilcox.test,
    map_signif_level = function(p) {
      if (p >= 0.001) {
        sprintf("p-value = %.3f", p)
      } else {
        "p < 0.001"
      }
    }
  ) +
  theme_minimal(base_size = 14) + # Slightly increase the base font size
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    # plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "Case Fatality Rate (CFR) Across Waves",
    x = "Wave",
    y = "Case Fatality Rate (per 100,000)"
  ) +
  annotate(
    "text",
    x = 0, y = 3200, # Adjust position based on your data range
    label = "Test: Wilcoxon Rank-Sum Test",
    hjust = 0, size = 3, color = "black", fontface = "italic"
  )

metric.dat %>%
  ggplot(aes(x = wave, y = HR, fill = wave)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(
      c("Wave 1", "Wave 2"),
      c("Wave 1", "Wave 3"),
      c("Wave 2", "Wave 3")
    ),
    y_position = c(16000, 18000, 20000), # Adjust for clear separation
    test = wilcox.test,
    map_signif_level = function(p) {
      if (p >= 0.001) {
        sprintf("p-value = %.3f", p)
      } else {
        "p < 0.001"
      }
    }
  ) +
  theme_minimal(base_size = 14) + # Slightly increase the base font size
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    # plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "Hospitalization Rate (HR) Across Waves",
    x = "Wave",
    y = "Hospitalization Rate (per 100,000)"
  ) +
  annotate(
    "text",
    x = 0, y = 21000, # Adjust position based on your data range
    label = "Test: Wilcoxon Rank-Sum Test",
    hjust = 0, size = 3, color = "black", fontface = "italic"
  )

metric.dat %>%
  ggplot(aes(x = wave, y = DP, fill = wave)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(
      c("Wave 1", "Wave 2"),
      c("Wave 1", "Wave 3"),
      c("Wave 2", "Wave 3")
    ),
    y_position = c(200, 220, 240), # Adjust for clear separation
    test = wilcox.test,
    map_signif_level = function(p) {
      if (p >= 0.001) {
        sprintf("p-value = %.3f", p)
      } else {
        "p < 0.001"
      }
    }
  ) +
  theme_minimal(base_size = 14) + # Slightly increase the base font size
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    # plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  labs(
    title = "Death per Population (DP) Across Waves",
    x = "Wave",
    y = "Death per Population (per 100,000)"
  ) +
  annotate(
    "text",
    x = 0, y = 250, # Adjust position based on your data range
    label = "Test: Wilcoxon Rank-Sum Test",
    hjust = 0, size = 3, color = "black", fontface = "italic"
  )

# kruskal.test(CFR ~ Wave, data = metric.dat)
# pairwise.wilcox.test(metric.dat$CFR, metric.dat$Wave,
#                  p.adjust.method = "BH")
```

```{r backup code}
# # Calculate CFR for each wave
# wave1_cfr <- dat |>
#   filter(date >= make_date(2020, 10, 1), date <= make_date(2021, 4, 1)) |>
#   group_by(state) |>
#   summarise(
#     cum_cases = sum(cases, na.rm = TRUE),
#     cum_deaths = sum(deaths, na.rm = TRUE),
#     cfr = ifelse(cum_cases > 0, cum_deaths / cum_cases * 100, NA)
#   )
# 
# wave2_cfr <- dat |>
#   filter(date >= make_date(2021, 7, 1), date <= make_date(2021, 12, 1)) |>
#   group_by(state) |>
#   summarise(
#     cum_cases = sum(cases, na.rm = TRUE),
#     cum_deaths = sum(deaths, na.rm = TRUE),
#     cfr = ifelse(cum_cases > 0, cum_deaths / cum_cases * 100, NA)
#   )
# 
# wave3_cfr <- dat |>
#   filter(date >= make_date(2021, 12, 2), date <= make_date(2022, 4, 1)) |>
#   group_by(state) |>
#   summarise(
#     cum_cases = sum(cases, na.rm = TRUE),
#     cum_deaths = sum(deaths, na.rm = TRUE),
#     cfr = ifelse(cum_cases > 0, cum_deaths / cum_cases * 100, NA)
#   )
# 
# # Combine CFR data
# wave1_cfr$wave <- "Wave 1"
# wave2_cfr$wave <- "Wave 2"
# wave3_cfr$wave <- "Wave 3"
# cfr_data <- bind_rows(wave1_cfr, wave2_cfr, wave3_cfr)
# 
# # Visualize CFR trends
# cfr_data |>
#   ggplot(aes(x = wave, y = cfr, fill = wave)) +
#   geom_boxplot(outlier.shape = NA) +
#   theme_minimal(base_size = 14) +
#   labs(
#     x = "Wave",
#     y = "Case Fatality Rate (CFR) (%)",
#     title = "Case Fatality Rate by Wave",
#     subtitle = "Comparison of CFR across 3 Waves"
#   )
```

## Regression (fitted) table of all states in the U.S.

```{r warning = FALSE}
##############################
# 7. Statistical Tests by State
##############################
all_states <- unique(analysis_data$state_name)

# Initialize a list to store the combined results for each state
all_state_results <- list()

for (st in all_states) {
  # Filter data for this state
  state_data <- analysis_data %>%
    filter(state_name == st)
  
  # Check if there's enough data for a model
  valid_points <- sum(!is.na(state_data$excess_deaths) & !is.na(state_data$covid_deaths))
  if (valid_points < 2) {
    # Not enough data to run a model, store placeholder
    all_state_results[[st]] <- data.frame(
      state_name = st,
      term = NA_character_,
      estimate = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value = NA_real_,
      r.squared = NA_real_
    )
    next
  }
  
  # Try fitting the model
  fit_result <- tryCatch({
    lm(excess_deaths ~ covid_deaths, data = state_data)
  }, error = function(e) NULL)
  
  if (is.null(fit_result)) {
    # Model failed, store placeholder
    all_state_results[[st]] <- data.frame(
      state_name = st,
      term = NA_character_,
      estimate = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value = NA_real_,
      r.squared = NA_real_
    )
    next
  }
  
  # Get coefficient details
  tidied <- broom::tidy(fit_result)
  # Get model-level statistics (including r.squared)
  glanced <- broom::glance(fit_result)
  
  # Add state_name and r.squared to each coefficient row
  tidied$state_name <- st
  tidied$r.squared <- glanced$r.squared
  
  all_state_results[[st]] <- tidied
}

# Combine all states into one data frame
final_results <- do.call(rbind, all_state_results)

# `final_results` now contains:
# state_name, term (Intercept or covid_deaths), estimate, std.error, statistic, p.value, and r.squared

final_results <- final_results %>%
  filter(!state_name %in% c("New York City", "United States"))
library(knitr)

final_results <- final_results %>%
  select(state_name, term, estimate, std.error, statistic, p.value, r.squared)

# Now neatly display the table
final_results %>%
  kable(digits = 3, caption = "Model Results by State") 

# Save plot to figs folder as a PNG
# ggsave(filename = "../output/excess_vs_covid_all_states.png", plot = p_all_states, width = 10, height = 8)
# write.csv(final_results, "../output/excess_mortality_models.csv", row.names = FALSE)
```