---
title: "BST 260 Final Project - Analysis"
author: "Weixuan Dong, Danny Lee, Bruce Zhou"
format:
  html:
    toc: TRUE
    toc_float: TRUE
    smooth-scroll: TRUE
---

# Load libraries

```{r}
require(tidyverse)
require(dplyr)
require(lubridate)
require(knitr)
require(ggsignif)
```

# Load data

```{r}
load('../data/dat.rda')
write_csv(dat, file = "../output/dat.csv")
```

# Analyses

## Pandemic period from Jan 2020 to Nov 2024

The figure below displays the COVID-19 cases, deaths, and hospitalization rate per 100,000 people in each region of the US from 2020 to 2024.

```{r trend plots}
# Define custom labels for the facets
dat.labs <- c("cases" = "COVID-19 Cases", "deaths" = "Deaths", "hosp" = "Hospitalizations")

# Apply to the ggplot code
dat %>%
  pivot_longer(cols = c("cases", "deaths", "hosp"), names_to = "outcome", values_to = "count") %>%
  mutate(rate = count * 100000 / population) %>%
  ggplot(aes(x = date, y = rate, group = state, color = region_name)) +
  geom_line(linewidth = 1) + # Thicker lines for better visibility
  facet_wrap(~outcome, ncol = 1, scales = "free", labeller = labeller(outcome = as_labeller(dat.labs))) +
  scale_x_date(
    breaks = "3 months", # Show a date label every 6 months
    date_labels = "%b %Y" # Format the labels as "Month Year" (e.g., Jan 2021)
  ) +
  theme_minimal(base_size = 14) + # Use a cleaner theme with larger base font
  theme(
    strip.text = element_text(face = "bold", size = 14), # Bold facet labels
    legend.position = "bottom", # Move legend to bottom for better space usage
    legend.title = element_text(face = "bold"), # Bold legend title
    axis.title = element_text(face = "bold"), # Bold axis titles
    axis.text.y = element_text(size = 10), # Slightly larger axis text for readability
    axis.text.x = element_text(size = 8, angle = 10),
    panel.grid.minor = element_blank() # Remove minor grid lines for a cleaner look
  ) +
  labs(
    x = "Date",
    y = "Rate (per 100,000 people)",
    color = "Region",
    title = "Rates of Cases, Deaths, and Hospitalizations Over Time (State Level)",
    subtitle = "Data displayed per 100,000 people by region"
  )

# ggsave(filename = "../output/covidwaves.pdf",plot = p1, dpi = 300, width = 14, height = 8, units = "in")
```

As can be seen from the trend plot which depicts the rates of COVID-19 cases, deaths, and hospitalizations from 01/25/2021 to 11/30/2024, there seems to be 3 waves of COVID-19 spread. The first one occurs from Oct 2020 to Apr 2021, where a peak in cases, deaths, and hospitalizations can be observed. Likewise, it appears that there was another wave between Jul 2021 to early Dec 2021. Lastly, there was a wave from early Dec 2021 to Apr 2022.

## Deaths rate by state during each time period

```{r}

dat %>%
  filter(date >= make_date(2020, 10, 1), date <= make_date(2021, 4, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  ) %>%
  ggplot(aes(x = reorder(state, death_rate), y = death_rate, fill = death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient(low = "#FFA5A5", high = "#AA0000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Death Rate (per 100,000)",
    fill = "Death Rate",
    title = "Cumulative COVID-19 Death Rates by State",
    subtitle = "Data from Oct 1st 2020 to Apr 1st 2021"
  )

dat %>%
  filter(date >= make_date(2021, 7, 1), date <= make_date(2021, 12, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  ) %>%
  ggplot(aes(x = reorder(state, death_rate), y = death_rate, fill = death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient(low = "#FFA5A5", high = "#AA0000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Death Rate (per 100,000)",
    fill = "Death Rate",
    title = "Cumulative COVID-19 Death Rates by State",
    subtitle = "Data from Jul 1st 2021 to Dec 1st 2021"
  )

dat %>%
  filter(date >= make_date(2021, 12, 2), date <= make_date(2022, 4, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  ) %>%
  ggplot(aes(x = reorder(state, death_rate), y = death_rate, fill = death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient(low = "#FFA5A5", high = "#AA0000") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Death Rate (per 100,000)",
    fill = "Death Rate",
    title = "Cumulative COVID-19 Death Rates by State",
    subtitle = "Data from Dec 2nd 2021 to Apr 1st 2022"
  )

wave1dat <- dat %>%
  filter(date >= make_date(2020, 10, 1), date <= make_date(2021, 4, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  )

wave2dat <- dat %>%
  filter(date >= make_date(2021, 7, 1), date <= make_date(2021, 12, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  )

wave3dat <- dat %>%
  filter(date >= make_date(2021, 12, 2), date <= make_date(2022, 4, 1)) %>%
  group_by(state) %>%
  summarise(
    cum_death = sum(deaths, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    death_rate = cum_death * 100000 / population
  )

# write_csv(wave1dat,file = "../output/wave1dat.csv")
# write_csv(wave2dat,file = "../output/wave2dat.csv")
# write_csv(wave3dat,file = "../output/wave3dat.csv")
```

From Oct 1st 2020 to Apr 1st 2021, overall this wave was arguably the most severe wave among all 3. The highest death rate per 100,000 people was reported in South Dakota (SD), with an average of 188.91 deaths among 100,000 people. Following South Dakota were North Dakota (177.97), Oklahoma (175.32), and Arkansas (150.17). There were also states which managed to contain the death rate below 50 per 100,000. For example, the lowest death rate was reported in Hawaii, which was only 9.12 deaths on average per 100,000 people, followed by Vermont (11.18), Alaska (26.48), Washington (39.58), Oregon (42.04), Puerto Rico (43.66), and Maine (47.19).

From July 1st 2021 to Dec 1st 2021, the COVID-19 variant seemed to be less virulent. The highest death rate per 100,000 people was reported in West Virginia (WV), with an average of 113.13 deaths among 100,000 people. Following West Virginia were Florida  (111.62), Montana (111.30), and Wyoming (109.88). There were also states which managed to contain the death rate below 40 per 100,000. For example, the lowest death rate was reported in Washington D.C., which was only 12.09 deaths on average per 100,000 people, followed by Vermont (12.39), Connecticut (12.39), New York (17.04), and Hawaii (36.83).

For the third wave, the COVID-19 variant during the period seemed to grow more infectious and deadly. The highest death rate per 100,000 people was reported in West Virginia (WV) again, with an average of 103.73 deaths among 100,000 people. Following West Virginia were Ohio (102.05), New Mexico (96.78), and Kentucky (93.82). There were also states which managed to contain the death rate below 40 per 100,000. For example, the lowest death rate was reported in Hawaii, which was only 20.76 deaths on average per 100,000 people, followed by Alaska (27.69), Puerto Rico (28.21), Vermont (29.85), and Utah (33.22).

```{r}
wave1dat$wave <- "Wave 1"
wave2dat$wave <- "Wave 2"
wave3dat$wave <- "Wave 3"
allWaves <- rbind(wave1dat, wave2dat, wave3dat)

allWaves %>%
  group_by(state) %>%
  summarise(avg_death_rate = mean(death_rate, na.rm = T)) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(state, avg_death_rate), y = avg_death_rate, fill = avg_death_rate)) + # Add `fill` aesthetic
  geom_col() +
  scale_fill_gradient2(low = "#A5D8FF", high = "#002D80") + # Define gradient colors
  coord_flip() +
  theme_minimal(base_size = 14) + # Minimal theme for better readability
  theme(
    axis.title.y = element_blank(), # Remove y-axis title for a cleaner look
    axis.title.x = element_text(face = "bold"), # Bold x-axis title
    axis.text = element_text(size = 6), # Slightly larger axis text
    legend.title = element_text(face = "bold"), # Bold legend title
    panel.grid.minor = element_blank(),
    legend.position = "right" # Position legend to the right
  ) +
  labs(
    y = "Average Death Rate (per 100,000)",
    fill = "Average Death Rate",
    title = "Average COVID-19 Death Rates by State across 3 Waves",
    subtitle = "Data from Oct 1st 2020 to Apr 1st 2022"
  )
```

We also compared the performance of each state across all 3 waves by taking the average of their death rate per 100,000 people. Results show that Oklahoma had the highest average death rate per 100,000, amounting to 120.00 deaths per 100,000 people, followed by West Virginia (115.33), Mississippi (109), Kentucky (107.42), Tennessee (106.92). There were also states which kept a relatively low death rate across waves. Namely, Vermont had an average death rate of only 17.81 per 100,000. Similarly, Hawaii had an average of 22.24 per 100,000, followed by Puerto Rico (30.94) New York (37.10), and Washington (40.07).

Generally, it can be seen that states in the northeastern part or the west coast of the US performed better than those in the middle or west US. It can also be observed that less populated states such as Vermont and Washington had lower death rates. But exceptions are also evident, for New York is rather populated but it was among the states with the lowest average death rates.

## COVID-19 became less or more virulent across different periods

To evaluate if COVID-19 became less or more virulent across different time periods, we will employ three key metrics: cases fatality rate (CFR), hospitalization rate (HR), and death per population (DP). The CFR is defined as deaths per case, HR is defined as hospitalizations per case, and DP is defined as deaths per population. In the following analyses, we will analyze these key metrics with figures and visualizations

### Distribution of the rates

```{r}
pooled.dat <- dat %>%
  mutate(Wave = case_when(
    date >= make_date(2020, 10, 1) & date <= make_date(2021, 4, 1) ~ "Wave 1",
    date >= make_date(2021, 7, 1) & date <= make_date(2021, 12, 1) ~ "Wave 2",
    date >= make_date(2021, 12, 2) & date <= make_date(2022, 4, 1) ~ "Wave 3",
    .default = NA
  )) %>%
  filter(!is.na(Wave)) %>%
  group_by(Wave, state) %>%
  summarise("Total Cases" = sum(cases, na.rm = T),
            "Total Deaths" = sum(deaths, na.rm = T),
            "Total Hospitalization" = sum(hosp, na.rm = T),
            "Total Population" = mean(population, na.rm = T)) %>%
  ungroup()

metric.dat <- pooled.dat %>%
  mutate(CFR = `Total Deaths` * 100000 / `Total Cases`,
         HR = `Total Hospitalization` * 100000 / `Total Cases`,
         DP = `Total Deaths` * 100000 / `Total Population`)


metric.dat %>%
  ggplot(aes(x = Wave, y = CFR, fill = Wave)) +
  geom_boxplot() +
  geom_signif(
    comparisons = list(c("Wave 1", "Wave 2"),
                       c("Wave 1", "Wave 3"),
                       c("Wave 2", "Wave 3")),
    y_position = c(2500, 2700, 3000)
  )

kruskal.test(CFR ~ Wave, data = metric.dat)
pairwise.wilcox.test(metric.dat$CFR, metric.dat$Wave,
                 p.adjust.method = "BH")
```


```{r}
# Define custom labels for the facets
dat.labs <- c("CFR" = "Cases Fatality Rate", "HR" = "Hospitalization Rate", "DP" = "Death per Population")

pooled.dat %>%
  group_by(Wave) %>%
  summarise("Total Cases" = sum(`Total Cases`, na.rm = T),
            "Total Deaths" = sum(`Total Deaths`, na.rm = T),
            "Total Hospitalization" = sum(`Total Hospitalization`, na.rm = T),
            "Total Population" = sum(`Total Population`, na.rm = T)) %>%
  ungroup() %>%
  mutate(CFR = `Total Deaths` * 100000 / `Total Cases`,
         HR = `Total Hospitalization` * 100000 / `Total Cases`,
         DP = `Total Deaths` * 100000 / `Total Population`) %>%
  kable("simple")

pooled.dat %>%
  group_by(Wave) %>%
  summarise("Total Cases" = sum(`Total Cases`, na.rm = T),
            "Total Deaths" = sum(`Total Deaths`, na.rm = T),
            "Total Hospitalization" = sum(`Total Hospitalization`, na.rm = T),
            "Total Population" = sum(`Total Population`, na.rm = T)) %>%
  ungroup() %>%
  mutate(CFR = `Total Deaths` * 100000 / `Total Cases`,
         HR = `Total Hospitalization` * 100000 / `Total Cases`,
         DP = `Total Deaths` * 100000 / `Total Population`) %>%
  pivot_longer(
    cols = c(CFR, HR, DP),
    names_to = "metric",
    values_to = "rate"
  ) %>%
  ggplot(aes(x = Wave, y = rate, fill = metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(
    ~metric, 
    scales = "free", 
    labeller = labeller(metric = dat.labs)
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold")
  ) +
  labs(
    title = "COVID-19 Metrics by Wave",
    x = "Wave",
    y = "Rate (per 100,000)"
  )
```

We also compared the average performance of each state across all 3 waves.

```{r}
# Calculate CFR for each wave
wave1_cfr <- dat |>
  filter(date >= make_date(2020, 10, 1), date <= make_date(2021, 4, 1)) |>
  group_by(state) |>
  summarise(
    cum_cases = sum(cases, na.rm = TRUE),
    cum_deaths = sum(deaths, na.rm = TRUE),
    cfr = ifelse(cum_cases > 0, cum_deaths / cum_cases * 100, NA)
  )

wave2_cfr <- dat |>
  filter(date >= make_date(2021, 7, 1), date <= make_date(2021, 12, 1)) |>
  group_by(state) |>
  summarise(
    cum_cases = sum(cases, na.rm = TRUE),
    cum_deaths = sum(deaths, na.rm = TRUE),
    cfr = ifelse(cum_cases > 0, cum_deaths / cum_cases * 100, NA)
  )

wave3_cfr <- dat |>
  filter(date >= make_date(2021, 12, 2), date <= make_date(2022, 4, 1)) |>
  group_by(state) |>
  summarise(
    cum_cases = sum(cases, na.rm = TRUE),
    cum_deaths = sum(deaths, na.rm = TRUE),
    cfr = ifelse(cum_cases > 0, cum_deaths / cum_cases * 100, NA)
  )

# Combine CFR data
wave1_cfr$wave <- "Wave 1"
wave2_cfr$wave <- "Wave 2"
wave3_cfr$wave <- "Wave 3"
cfr_data <- bind_rows(wave1_cfr, wave2_cfr, wave3_cfr)

# Visualize CFR trends
cfr_data |>
  ggplot(aes(x = wave, y = cfr, fill = wave)) +
  geom_boxplot(outlier.shape = NA) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Wave",
    y = "Case Fatality Rate (CFR) (%)",
    title = "Case Fatality Rate by Wave",
    subtitle = "Comparison of CFR across 3 Waves"
  )
```

To evaluate virulence, we calculate Case Fatality Rate (CFR) across waves. CFR is calculated as $CFR = \frac{Cumulative Deaths}{Cumulative Cases} \cdot 100$. 
We can clearly see from the plot that COVID-19 became less virulent across the three waves, as evidenced by the declining CFR from Wave 1 to Wave 3. While the initial wave saw a relatively high median CFR around 1.5%, this dropped progressively in subsequent waves, reaching below 1% by Wave 3. The reduction in fatality rates is likely due to increased vaccination coverage, improved medical responses, and healthcare system adaptations.

```{r}
# Aggregate data to weekly level
weekly_data <- dat |>
  mutate(week = floor_date(date, "week")) |>
  group_by(state, week) |>
  summarise(
    observed_deaths = sum(deaths, na.rm = TRUE),
    covid_deaths = sum(deaths, na.rm = TRUE) # Assuming "deaths" include COVID deaths
  )

# Use 2024 data to calculate the baseline expected deaths
baseline <- weekly_data |>
  filter(week >= make_date(2024, 1, 1), week <= make_date(2024, 12, 31)) |>
  group_by(state) |>
  summarise(expected = mean(observed_deaths, na.rm = TRUE))

# Merge baseline expected deaths into the full dataset
excess_mortality <- weekly_data |>
  left_join(baseline, by = "state") |>
  mutate(
    excess_deaths = observed_deaths - expected,
    covid_excess_ratio = ifelse(excess_deaths > 0, covid_deaths / excess_deaths, NA)
  )

# Visualize excess mortality by week and state
excess_mortality |>
  ggplot(aes(x = week, y = excess_deaths, color = state)) +
  geom_line() +
  theme_minimal(base_size = 14) +
  labs(
    x = "Week",
    y = "Excess Mortality",
    title = "Weekly Excess Mortality by State",
    subtitle = "Comparison of Excess Mortality and COVID-19 Deaths"
  )

# Save results
write_csv(excess_mortality, file = "../output/excess_mortality.csv")
```

We used the observed weekly deaths from 2024 to calculate the average weekly deaths for each state. This serves as a proxy for "expected" deaths in a stable post-pandemic period. Then we compared the observed deaths during the pandemic to the expected deaths derived from 2024 data by plotting the excess mortality.

According to the line plot showing the weekly trends in excess mortality across states, we can see that there has been excess deaths during 2020 - 2022, where COVID-19 was severely affecting the death rate. Thus, we can conclude that COVID-19 deaths explain the excess mortality fairly well, assuming that 2024 (post-pandemic baseline) represents a "normalized" year.
